---
title: "using custom rockfish database to assign ASVs to taxa"
author: "Kimberly Ledger"
date: "2024-11-06"
output: html_document
---

```{r load-libraries}
library(tidyverse)
rename <- dplyr::rename
```

input blastn results from custom rockfish database
```{r}
#rock <- read.table("/home/kimberly.ledger/rockfish_mb/data/blastn_tissue_mock_bt22_db536.txt")  %>%
#rock <- read.table("/home/kimberly.ledger/rockfish_mb/data/blastn_tissue_mock_bt22_db534.txt")  %>%
rock <- read.table("/home/kimberly.ledger/rockfish_mb/data/blastn_tissue_mock_bt22_534_20250117.txt")  %>%
    filter(V4 > 100) %>% # make sure all retained matches are >100 bp
  select(-V5, -V6, -V7, -V8, -V9, -V10, -V11, -V12, -V13, -V14) %>%  #remove unnecessary columns
  group_by(V1, V2) %>% # group by the sequence key and the full taxonomy to reduce duplicate entries
  unique() %>% # doing that reduced the number of entries
  rename(qseqid=V1, accession_id=V2, perc_id=V3, length=V4) #%>% #rename headers
  #filter(perc_id > 98)
```

load the id's of accessions
```{r}
#id <- read.csv("/home/kimberly.ledger/rockfish_mb/data/rkfish_ref_dbs/rockfish_reference_db_536.csv") %>%
#id <- read.csv("/home/kimberly.ledger/rockfish_mb/data/rkfish_ref_dbs/rockfish_reference_db_534.csv") %>%
id <- read.csv("/home/kimberly.ledger/rockfish_mb/data/rkfish_ref_dbs/rockfish_reference_db_534_20250117.csv") %>%
  rename(accession_id = Name)

id %>% select(Organism) %>% unique()
```

join 
```{r}
rock_id <- rock %>%
  left_join(id) %>%
  rename(species = Organism)
```

## rank the hits by % identity
```{r}
seq_id_diff <- rock_id %>%
  select(-length) %>%
  group_by(qseqid, species) %>%
    mutate(seq_percID = max(perc_id)) %>%
    group_by(qseqid, species, seq_percID) %>%
  summarise(max(seq_percID)) %>% # take just the top hit for each taxon (for each sequence)
  select(-`max(seq_percID)`) %>%
  ungroup() %>%
  group_by(qseqid) %>%
      #mutate(id_rank = rank(desc(seq_percID), ties.method = "min")) %>% # rank the taxonomic hits per sequence by % id
       mutate(top_perc = max(seq_percID)) %>% # designate the highest % id for the best taxonomic hit in each sequence (in some, but not all cases, this is 100%)   %>%
      mutate(diff = top_perc - seq_percID) %>% # calculate the difference between the % identity of the top hit and each subsequent taxonomic hit
      arrange(diff)

seq_id_diff
```


since there are very few bp difference among species, let's go with species within 0.5% of the top match and top match must be at least 95% 
```{r}
keepers <- seq_id_diff %>%
  filter(diff < 0.5) %>%   #the 0.5% threshold essentially is requiring a 100% match 
  filter(top_perc > 95)
```


## so now determine whether ASVs should be assigned to genus, family, order, etc. 
```{r}
singletons <- keepers %>%
  select(qseqid) %>%
  tally() %>%
  filter(n == 1)

# these are the seqs that now have only a single match
singleton_df <- singletons %>%
  left_join(keepers) %>%
  select(-n) %>%
  mutate(taxonomic_level = "species") %>%
  mutate(taxon = species) #%>%
  #select(qseqid, taxon, taxonomic_level)

## Genus-level matches
# remove the singletons from the bigger df 
single_genus <- keepers %>%
  anti_join(singleton_df)%>% # at best, these should be genus-level matches
  left_join(., keepers) %>%
  arrange(qseqid) %>%
  mutate(taxonomic_level = "genus") %>%
  mutate(taxon = "Sebastes") #%>%
  #select(qseqid, taxon, taxonomic_level) %>%
  #unique
```

change non-species level ASV assignments so that it represents the species it makes up. 
- Sebastes_ev = Sebastes emphaeus or Sebastes variegatus
- Sebastes_pv  = 	Sebastes polyspinis or Sebastes variabilis
- etc.... 

```{r}
single_genus <- single_genus %>%
  group_by(qseqid) %>%
  mutate(taxon = case_when(
    any(species == "Sebastes ciliatus") & any(species == "Sebastes ciliatus_variabilis") ~ "Sebastes ciliatus_variabilis",
    TRUE ~ taxon
  )) %>%
  ungroup()

single_genus <- single_genus %>%
  group_by(qseqid) %>%
  mutate(taxon = case_when(
    any(species == "Sebastes variabilis") & any(species == "Sebastes ciliatus_variabilis") ~ "Sebastes ciliatus_variabilis",
    TRUE ~ taxon
  )) %>%
  ungroup()

single_genus <- single_genus %>%
  group_by(qseqid) %>%
  mutate(taxon = case_when(
    any(species == "Sebastes variegatus") & any(species == "Sebastes wilsoni") ~ "Sebastes vw",
    TRUE ~ taxon
  )) %>%
  ungroup()

single_genus <- single_genus %>%
  group_by(qseqid) %>%
  mutate(taxon = case_when(
    any(species == "Sebastes emphaeus") & any(species == "Sebastes variegatus") ~ "Sebastes ev",
    TRUE ~ taxon
  )) %>%
  ungroup()

single_genus <- single_genus %>%
  group_by(qseqid) %>%
  mutate(taxon = case_when(
    any(species == "Sebastes diaconus") & any(species == "Sebastes entomelas") ~ "Sebastes de",
    TRUE ~ taxon
  )) %>%
  ungroup()

single_genus <- single_genus %>%
  group_by(qseqid) %>%
  mutate(taxon = case_when(
    any(species == "Sebastes polyspinis") & any(species == "Sebastes ciliatus_variabilis") ~ "Sebastes ciliatus_variabilis_polyspinus",
    TRUE ~ taxon
  )) %>%
  ungroup()

# single_genus <- single_genus %>%
#   group_by(qseqid) %>%
#   mutate(taxon = case_when(
#     any(species == "Sebastes reedi") & any(species == "Sebastes crameri") ~ "Sebastes rc",
#     TRUE ~ taxon
#   )) %>%
#   ungroup()
# 
# single_genus <- single_genus %>%
#   group_by(qseqid) %>%
#   mutate(taxon = case_when(
#     any(species == "Sebastes entomelas") & any(species == "Sebastes mystinus") ~ "Sebastes em",
#     TRUE ~ taxon
#   )) %>%
#   ungroup()
# 
# single_genus <- single_genus %>%
#   group_by(qseqid) %>%
#   mutate(taxon = case_when(
#     any(species == "Sebastes crocotulus") & any(species == "Sebastes miniatus") ~ "Sebastes cm",
#     TRUE ~ taxon
#   )) %>%
#   ungroup()
# 
# single_genus <- single_genus %>%
#   group_by(qseqid) %>%
#   mutate(taxon = case_when(
#     any(species == "Sebastes ciliatus") & any(species == "Sebastes polyspinis") ~ "Sebastes cp",
#     TRUE ~ taxon
#   )) %>%
#   ungroup()
# 
# 
# 
# single_genus <- single_genus %>%
#   group_by(qseqid) %>%
#   mutate(taxon = case_when(
#     any(species == "Sebastes chlorostictus") & any(species == "Sebastes rosenblatti") ~ "Sebastes cr",
#     TRUE ~ taxon
#   )) %>%
#   ungroup()
# 
# single_genus <- single_genus %>%
#   group_by(qseqid) %>%
#   mutate(taxon = case_when(
#     any(species == "Sebastes variegatus") & any(species == "Sebastes wilsoni") & any(species == "Sebastes zacentrus") ~ "Sebastes vwz",
#     TRUE ~ taxon
#   )) %>%
#   ungroup()
# 
# single_genus <- single_genus %>%
#   group_by(qseqid) %>%
#   mutate(taxon = case_when(
#     any(species == "Sebastes ciliatus") & any(species == "Sebastes polyspinis") & any(species == "Sebastes variabilis")  ~ "Sebastes cpv",
#     TRUE ~ taxon
#   )) %>%
#   ungroup()
```

```{r}
rockfish_df <- singleton_df %>%
  bind_rows(single_genus)
```

```{r}
rockfish_df_collapsed <- rockfish_df  %>%
   select(qseqid, taxon) %>%
   unique()
```

```{r}
#write.csv(rockfish_df_collapsed, "/home/kimberly.ledger/rockfish_mb/data/taxonomy_custom534_20250117.csv")
```



#### an older aside that just looked at assignment of dusky, dark, and northern 

just looking at the dark, northern, and dusky reference db results 

first figure out which asv's got id'd to one of these species in the original assignment 
```{r}
cpv_asvs <- read.csv("/home/kimberly.ledger/rockfish_mb/data/taxonomy_custom534_20250107.csv") %>%
  select(!X) %>%
  filter(taxon %in% c("Sebastes ciliatus", "Sebastes polyspinis", "Sebastes variabilis"))
```


input blastn results from custom rockfish database
```{r}
#rock <- read.table("/home/kimberly.ledger/rockfish_mb/data/blastn_tissue_mock_bt22_db536.txt")  %>%
rock <- read.table("/home/kimberly.ledger/rockfish_mb/data/blastn_tissue_mock_bt22_cpv.txt")  %>%
    filter(V4 > 100) %>% # make sure all retained matches are >100 bp
  select(-V5, -V6, -V7, -V8, -V9, -V10, -V11, -V12, -V13, -V14) %>%  #remove unnecessary columns
  group_by(V1, V2) %>% # group by the sequence key and the full taxonomy to reduce duplicate entries
  unique() %>% # doing that reduced the number of entries
  rename(qseqid=V1, accession_id=V2, perc_id=V3, length=V4) #%>% #rename headers
  #filter(perc_id > 98)
```

load the id's of accessions
```{r}
#id <- read.csv("/home/kimberly.ledger/rockfish_mb/data/rkfish_ref_dbs/rockfish_reference_db_536.csv") %>%
id <- read.csv("/home/kimberly.ledger/rockfish_mb/data/rkfish_ref_dbs/cpv_minidloop_representative_haplotypes.csv") %>%
  rename(accession_id = Name)

id %>% select(Organism) %>% unique()
```



join 
```{r}
rock_id <- rock %>%
  left_join(id) %>%
  rename(species = Organism)
```

## rank the hits by % identity
```{r}
seq_id_diff <- rock_id %>%
  select(-length) %>%
  group_by(qseqid, accession_id) %>%
    mutate(seq_percID = max(perc_id)) %>%
    group_by(qseqid, accession_id, seq_percID) %>%
  summarise(max(seq_percID)) %>% # take just the top hit for each taxon (for each sequence)
  select(-`max(seq_percID)`) %>%
  ungroup() %>%
  group_by(qseqid) %>%
      #mutate(id_rank = rank(desc(seq_percID), ties.method = "min")) %>% # rank the taxonomic hits per sequence by % id
       mutate(top_perc = max(seq_percID)) %>% # designate the highest % id for the best taxonomic hit in each sequence (in some, but not all cases, this is 100%)   %>%
      mutate(diff = top_perc - seq_percID) %>% # calculate the difference between the % identity of the top hit and each subsequent taxonomic hit
      arrange(diff)

seq_id_diff
```


since there are very few bp difference among species, let's go with species within 0.5% of the top match and top match must be at least 95% 
```{r}
keepers <- seq_id_diff %>%
  filter(diff < 0.5) %>%   #the 0.5% threshold essentially is requiring a 100% match 
  filter(top_perc > 95)
```

how do these keeps match the orginal assignments 

```{r}
temp <- keepers %>%
  left_join(cpv_asvs)
```

okay, based on this i'll only keep the 98% matches or better 
```{r}
keepers <- keepers %>%
  filter(top_perc > 98)
```

so haps 1, and 7 thru 12 are likely polyspinis
and haps 2-6 are a toss up between ciliatus and variabilis 

```{r}
keepers_w_id <- keepers %>%
  separate(accession_id, into = c("hap")) %>%
  mutate(taxon = ifelse(hap == "hap2", "Sebastes polyspinis", NA),
         taxon = ifelse(hap == "hap3", "Sebastes polyspinis", taxon),
         taxon = ifelse(hap == "hap4", "Sebastes polyspinis", taxon),
         taxon = ifelse(hap == "hap5", "Sebastes polyspinis", taxon),
         taxon = ifelse(hap == "hap6", "Sebastes polyspinis", taxon),
         taxon = ifelse(hap == "hap1", "Sebastes ciliatus/variabilis", taxon),
         taxon = ifelse(hap == "hap7", "Sebastes ciliatus/variabilis", taxon),
         taxon = ifelse(hap == "hap8", "Sebastes ciliatus/variabilis", taxon),
         taxon = ifelse(hap == "hap9", "Sebastes ciliatus/variabilis", taxon),
         taxon = ifelse(hap == "hap10", "Sebastes ciliatus/variabilis", taxon),
         taxon = ifelse(hap == "hap11", "Sebastes ciliatus/variabilis", taxon),
         taxon = ifelse(hap == "hap12", "Sebastes ciliatus/variabilis", taxon))
```


## so now determine whether ASVs should be assigned to genus, family, order, etc. 
```{r}
singletons <- keepers_w_id %>%
  select(qseqid) %>%
  tally() %>%
  filter(n == 1)

# these are the seqs that now have only a single match
singleton_df <- singletons %>%
  left_join(keepers_w_id) %>%
  select(-n)

## Genus-level matches
# remove the singletons from the bigger df 
not_singleton <- keepers_w_id %>%
  anti_join(singleton_df)%>% # at best, these should be genus-level matches
  left_join(., keepers_w_id) %>%
  arrange(qseqid) 
```
only ASV 199 is stuck between the northerns and dusky/dark groups - for simpicity for now i will call them northern 

```{r}
not_singleton <- not_singleton %>%
  filter(taxon != "Sebastes ciliatus/variabilis") %>%
  select(!hap) %>%
  unique()
```

```{r}
cpv_df <- singleton_df %>%
  bind_rows(not_singleton)
```


```{r}
write.csv(cpv_df, "/home/kimberly.ledger/rockfish_mb/data/taxonomy_cpv_only.csv")
```