---
title: "rockfish eDNA samples"
output: html_document
date: "2024-12-20"
---

libraries
```{r}
library(tidyverse)
rename <- dplyr::rename
library(reshape2)
```

load sample metadata
```{r}
metadata <- read.csv("/home/kimberly.ledger/rockfish_mb/data/metadata_tissue_mock_bt22.csv") %>%
  rename(Sample_ID = sample_ID)
```

load decontaminated reads 
```{r}
asv_table <- read.csv("/home/kimberly.ledger/rockfish_mb/data/decontaminated_asv_table.csv") %>%   ##this asv does not include reps w/ less than 200 reads
  select(!X) %>%
  filter(!sample_type == "tissue") %>%
  filter(!sample_type == "mock") %>%
  filter(Sample_ID != "NC") %>%
  filter(Sample_ID != "PC") %>%
  filter(Sample_ID != "NTC") %>%  ### these were all controls associated with tissue or mock amplification 
  #select(!sample_type) %>% 
  filter(reads > 0)
```

convert asv table to taxon table
```{r}
taxonomy <- tax <- read.csv("/home/kimberly.ledger/rockfish_mb/data/taxonomy_custom534_20250117.csv") %>% 
  select(!X) %>%
  select(qseqid, taxon) %>%
  unique() %>%
  rename(ASV = qseqid) 

taxon_table <- asv_table %>%
  left_join(taxonomy, by="ASV") %>%
  group_by(Sample_ID, sample_type, taxon) %>%
  summarize(tot_reads = sum(reads)) %>%
  filter(tot_reads > 0) 
```


let me see what the aquarium samples look like 
```{r}
tank_fig1 <- taxon_table %>%
  filter(Sample_ID %in% c("e01737", "e01738", "e01741")) %>%
  group_by(Sample_ID) %>%
  mutate(reads_per_sample = sum(tot_reads)) %>%
  mutate(prop_reads = tot_reads/reads_per_sample) %>%
  ggplot(aes(x=Sample_ID, y=prop_reads, fill=taxon)) +
  geom_bar(stat = "identity") + 
  theme_bw() +
  labs(
    y = "relative read abundance (%)",
    x = "sample_id") + 
  theme(#axis.text.x = element_blank(),
    axis.text.x = element_text(angle = 90, hjust = 0.95),
    legend.text = element_text(size = 8),
    legend.key.size = unit(0.3, "cm"),
    legend.position = "bottom",
    legend.title = element_blank()
  )

tank_fig1

tank_table <- taxon_table %>%
  filter(Sample_ID %in% c("e01737", "e01738", "e01741")) %>%
  group_by(Sample_ID) %>%
  mutate(reads_per_sample = sum(tot_reads)) %>%
  mutate(prop_reads = tot_reads/reads_per_sample)
```

```{r}
ggsave("/home/kimberly.ledger/rockfish_mb/figures/tank1.png", plot = tank_fig1, dpi = 300, width =6, height = 4)
```


make a heatmap table for pres/abs of replicates 
```{r}
tank_pres <- tank_table %>%
  select(Sample_ID, taxon, prop_reads) %>%
  pivot_wider(names_from = taxon, values_from = prop_reads, values_fill = 0) %>%
  pivot_longer(cols = !Sample_ID, names_to = "taxon", values_to = "prop_reads") %>%
  mutate(presence = ifelse(prop_reads > 0, 1, 0)) %>%
  mutate(Sample_ID = ifelse(Sample_ID == "e01737", "A", Sample_ID), 
         Sample_ID = ifelse(Sample_ID == "e01738", "B", Sample_ID),
         Sample_ID = ifelse(Sample_ID == "e01741", "C", Sample_ID))

tank_pres$taxon <- factor(tank_pres$taxon, levels = rev(sort(unique(tank_pres$taxon))))

tank_pres_plot <- ggplot(tank_pres, aes(x = Sample_ID, y = taxon, fill = as.factor(presence))) +
  geom_tile(color = "white") +
  scale_fill_manual(values = c("0" = "white", "1" = "grey"), 
                    name = "", 
                    labels = c("Absent", "Present")) +
  theme_minimal() +
  labs(#title = "taxon Presence Across Replicates",
       x = "replicate", 
       y = "") +
  theme(axis.text.x = element_text(angle = 0, hjust = 1))
tank_pres_plot 
```

```{r}
ggsave("/home/kimberly.ledger/rockfish_mb/figures/tank4_table.png", plot = tank_pres_plot, dpi = 300, width = 4, height = 4)
```


and a quick look a field samples 
```{r}
field_table <- taxon_table %>%
  filter(sample_type == "sample") %>%
  filter(!Sample_ID %in% c("e01737", "e01738", "e01741")) %>% 
  filter(tot_reads > 0) %>%
  group_by(taxon) %>%
  mutate(total_taxon_reads = sum(tot_reads)) %>%
  ungroup() %>% 
  filter(total_taxon_reads > 30) %>%    ## remove taxa with <30 reads total 
  group_by(Sample_ID) %>%
  mutate(reads_per_sample = sum(tot_reads)) %>%
  mutate(prop_reads = tot_reads/reads_per_sample)

bt_plot1 <- field_table %>%
  ggplot(aes(x=Sample_ID, y=prop_reads, fill=taxon)) +
  geom_bar(stat = "identity") + 
  theme_bw() +
  labs(
    y = "relative read abundance (%)",
    x = "sample_id") + 
  theme(axis.text.x = element_blank(),
    #axis.text.x = element_text(angle = 90, hjust = 0.95),
    legend.text = element_text(size = 8),
    legend.key.size = unit(0.3, "cm"),
    legend.position = "bottom",
    legend.title = element_blank()
  )

bt_plot1

field_table %>%
  group_by(taxon) %>%
  summarize(reads = sum(tot_reads)) %>%
  arrange(desc(reads))
```

mostly POP and northern in BT field samples. that makes sense!  

let me add in some field metadata about location - join to the full field metadata to not lose the 0 samples 
```{r}
field_table_w_meta <- metadata %>%
  filter(sample_type == "sample") %>%
  filter(!Sample_ID %in% c("e01737", "e01738", "e01741")) %>% 
  left_join(field_table, by = "Sample_ID") %>%
  mutate(prop_reads = ifelse(is.na(prop_reads), 0, prop_reads)) %>%
  filter(!Sample_ID %in% c("e02339", "e02261_A", "e02298"))
```

```{r}
field_table_w_meta$station <- as.numeric(field_table_w_meta$station)
field_table_w_meta$station <- factor(field_table_w_meta$station, levels = sort(unique(field_table_w_meta$station), decreasing = TRUE))

field_table_w_meta %>%
  select(station, extraction_ID) %>%
  unique() %>%
  group_by(station) %>%
  summarise(n = n())

my_colors <- c("#f5a056",  "#6e9bf4", "#cc7af4", "#1f77b4", "#77ab31")

field_table_w_meta <- field_table_w_meta %>%
  mutate(taxon_abbr = sub("^(\\w)\\w*\\s(\\w+)", "\\1. \\2", taxon))
  #mutate(taxon_abbr = ifelse(taxon == "Sebastes ciliatus_variabilis_polyspinus", "Sebastes cvp", taxon), 
         #taxon_abbr = ifelse(taxon == "Sebastes ciliatus_variabilis", "Sebastes cv", taxon_abbr))

bt_plot2 <- field_table_w_meta %>%
  mutate(station_dummy = as.numeric(station)) %>%
  ggplot(aes(x=Sample_ID, y=prop_reads, fill=taxon_abbr)) +
  geom_bar(stat = "identity") + 
  theme_bw() +
  scale_fill_manual(values = my_colors, na.translate = FALSE) + # Removes NA from the legend
  facet_grid(~station_dummy, scales = "free_x") +
  labs(
    y = "relative read abundance (%)",
    x = "samples by haul") + 
  theme_bw() +
  theme(axis.text.x = element_blank(),
    #axis.text.x = element_text(angle = 90, hjust = 0.95),
    legend.text = element_text(size = 8),
    legend.key.size = unit(0.3, "cm"),
    legend.position = "bottom",
    legend.title = element_blank(),
    #strip.text = element_blank(), # Removes facet labels
    axis.ticks.x = element_blank() 
  )
  
bt_plot2
```

```{r}
ggsave("/home/kimberly.ledger/rockfish_mb/figures/bt2_w_zeros_dummyfacets.png", plot = bt_plot2, dpi = 300, width = 8, height =3)
```

```{r}
write.csv(field_table_w_meta, "/home/kimberly.ledger/rockfish_mb/outputs/field_quant/raw_field_data.csv")
```


```{r}
field_table_w_meta %>%
  filter(is.na(taxon))

field_table_w_meta %>%
  select(Sample_ID) %>%
  unique()

field_table_w_meta %>%
  select(Sample_ID, reads_per_sample) %>%
  unique() %>%
  filter(!is.na(reads_per_sample)) %>%
  summarize(mean = mean(reads_per_sample),
            sd = sd(reads_per_sample))

field_table_w_meta %>%
  group_by(taxon) %>%
  filter(tot_reads > 100) %>%
  summarize(count = n())
```

controls 
```{r}
controls_table <- taxon_table %>%
  filter(sample_type != "sample") %>%
  group_by(Sample_ID) %>%
   mutate(reads_per_sample = sum(tot_reads)) %>%
   mutate(prop_reads = tot_reads/reads_per_sample) %>%
  filter(sample_type == "field_blank") %>%
  filter(prop_reads > 0.25)
```

check how many eDNA samples were collected at each haul location 
```{r}
bt_meta <- metadata %>%
  filter(sample_type == "sample") %>%
  filter(!Sample_ID %in% c("e01737", "e01738", "e01741")) %>%
  group_by(station) %>%
  summarize(count = n())

metadata %>%
  filter(sample_type == "sample") %>%
  filter(!Sample_ID %in% c("e01737", "e01738", "e01741")) %>%
  select(station, depth) %>%
  unique()
```

#### the following code is not used in the rockfish publication ####

how does this compare to trawl catch? 

load Alaska Groundfish Survey catch and haul report for fish from the 2022 Aleutian Island BT survey: https://www.fisheries.noaa.gov/foss/f?p=215:28:51549328646:::::
```{r}
bt_catch <- read.csv("/home/kimberly.ledger/BT2021_metabarcoding/data/2022_AI_BT_Teleostei_CATCH_AND_HAUL_DATA_TAXON.csv") %>%
  filter(Vessel.name == "OCEAN EXPLORER") %>%
  filter(Haul.number %in% field_table_w_meta$station)
```

metadata from bt hauls w/ eDNA samples
```{r}
bt_catch %>%
  select(Survey.year:Surface.temperature..degrees.Celsius.) %>%
  unique() %>%  
  select(Stratum.ID, Station.ID, Haul.ID, Haul.number, Date.time:Surface.temperature..degrees.Celsius.) %>%  ## just the columns that vary
  select(Haul.number, Depth..m.)
```

good, this matches the metadata i have from the eDNA samples. 

```{r}
bt_catch %>%
  select(Haul.number, Taxon.common.name:Taxon.confidence.rating) %>%
  arrange(Haul.number)
```

```{r}
bt_sebastes <- bt_catch %>%
  select(Haul.number, Taxon.common.name:Taxon.confidence.rating) %>%
  arrange(Haul.number) %>%
  filter(str_detect(Taxon.scientific.name, "Sebastes"))

#bt_sebastolobus <- bt_catch %>%
#  select(Haul.number, Taxon.common.name:Taxon.confidence.rating) %>%
#  arrange(Haul.number) %>%
#  filter(str_detect(Taxon.scientific.name, "Sebastolobus"))
```

only northerns and one pop in hauls associated with eDNA samples 

how about what sebastes taxon were caught in the entire 2022 AI bottom trawl survey? 
```{r}
seb_in_bt <- read.csv("/home/kimberly.ledger/BT2021_metabarcoding/data/2022_AI_BT_Teleostei_CATCH_AND_HAUL_DATA_TAXON.csv") %>%
  filter(Vessel.name == "OCEAN EXPLORER") %>% 
  filter(str_detect(Taxon.scientific.name, "Sebast"))

unique(seb_in_bt$Taxon.scientific.name)
```

```{r}
seb_in_bt %>%
  group_by(Taxon.scientific.name) %>%
  summarize(tot_wt_CPUE = sum(Weight.CPUE..kg.km2.),
            tot_n_CPUE = sum(Number.CPUE..no.km2.)) %>%
  arrange(desc(tot_wt_CPUE))

seb_in_bt %>% select(Haul.number) %>% unique()

seb_in_bt %>% 
  filter(Haul.number %in% c("107", "114", "129", "125", "138", "152", "90", "92", "95", "99")) %>%
  group_by(Taxon.scientific.name) %>%
  summarize(tot_wt_CPUE = sum(Weight.CPUE..kg.km2.),
            tot_n_CPUE = sum(Number.CPUE..no.km2.)) %>%
  arrange(desc(tot_wt_CPUE))
```

try making a map of these trawl locations
```{r}
library(akgfmaps)
browseVignettes('akgfmaps')

?make_idw_map
?get_base_layers

ai.central <- get_base_layers(select.region = "ai.central", set.crs = 3338)

ai.central_plot <- ggplot() +
  #geom_sf(data = ai.central$bathymetry) +
  geom_sf(data = ai.central$survey.area) +
  geom_sf(data = ai.central$akland, fill = "grey70") +
  coord_sf(xlim = ai.central$plot.boundary$x,
           ylim = ai.central$plot.boundary$y) +
  theme_bw()
ai.central_plot

ai <- get_base_layers(select.region = "ai", set.crs = 3338)

ai_plot <- ggplot() +
  #geom_sf(data = ai$bathymetry) +
  geom_sf(data = ai$survey.area) +
  geom_sf(data = ai$akland, fill = "grey70") +
  coord_sf(xlim = ai$plot.boundary$x,
           ylim = ai$plot.boundary$y) +
  theme_bw()
ai_plot
```

```{r}
#sites_2 <- field_table_w_meta %>%
#  ungroup() %>%
#  select(station, longitude, latitude, depth) %>%
#  unique()

sites <- bt_catch %>%
  rename(latitude = Start.latitude..decimal.degrees.) %>%
  rename(longitude = Start.longitude..decimal.degrees.) %>%
  select(Haul.number, longitude, latitude) %>%
  unique()

# Step 1: Convert to sf object (default CRS is WGS84, EPSG:4326)
sites_sf <- st_as_sf(sites, coords = c("longitude", "latitude"), crs = 4326)

# Step 2: Transform to Alaska Albers Equal Area (EPSG:3338)
sites_sf_3338 <- st_transform(sites_sf, crs = 3338)

# Print the resulting spatial object
print(sites_sf_3338)

# Optional: Plot the points on a map
library(ggplot2)
ggplot() +
  geom_sf(data = sites_sf_3338) +
  theme_minimal() +
  ggtitle("Points in Alaska Albers Equal Area Projection (EPSG:3338)")
```

```{r}
ai_plot <- ggplot() +
  #geom_sf(data = ai$bathymetry) +
  geom_sf(data = ai$survey.area) +
  geom_sf(data = ai$akland, fill = "grey70") +
  geom_sf(data = sites_sf_3338) +
  coord_sf(xlim = ai$plot.boundary$x,
           ylim = ai$plot.boundary$y) +
  theme_bw()
ai_plot
```



```{r}
taxon_table %>%
  filter(sample_type %in% c("extraction_blank", "field_blank", "pcr_blank", "positive")) %>%
  group_by(Sample_ID) %>%
  mutate(reads_per_sample = sum(tot_reads)) %>%
  mutate(prop_reads = tot_reads/reads_per_sample) %>%
  #filter(reads_per_sample > 100) %>%
  ggplot(aes(x=Sample_ID, y=tot_reads, fill=taxon)) +
  #ggplot(aes(x=Sample_ID, y=prop_reads, fill=taxon)) +
  geom_bar(stat = "identity") + 
  theme_bw() +
  facet_wrap(~sample_type, scales = "free_x") + 
  labs(
    y = "relative read abundance (%)",
    x = "sample_id") + 
  theme(#axis.text.x = element_blank(),
    axis.text.x = element_text(angle = 90, hjust = 0.95),
    legend.text = element_text(size = 8),
    legend.key.size = unit(0.3, "cm"),
    legend.position = "bottom",
    legend.title = element_blank()
  )
```

i flipped the label for the NC/PC samples (and they were POP) 
i used gopher rf for PC22_A 

```{r}
asv_table_with_sample_type %>%
  filter(Sample_ID == "PC22_A") %>%
  pivot_longer(cols = c(asv_first:asv_last), names_to = "ASV", values_to = "reads") %>%
  filter(reads > 0)
```

