---
title: "add custom ref seqs"
output: html_document
date: "2024-12-05"
---

this code takes the asvs from the rockfish mb tissues and creates ref sequences for the few species we are missing/low of reference sequences 

```{r}
library(tidyverse)
```

load asv table
```{r}
asv_table <- read.csv("/genetics/edna/workdir/sebastes/tissues/untrimmed_20241231/filtered/outputs/ASVtable.csv") %>%
  dplyr::rename(sample_id = X)
```

read in file that has the tissue ids for the rockfish samples 
```{r}
ablg <- read.csv("~/rockfish_mb/data/ABLG_BURKE_rockfishDNA.csv")
```

join
```{r}
asv_table <- asv_table %>%
  left_join(ablg, by = "sample_id") %>%
  select(sample_id, common_name, species_name, everything())
```


#DEACON

pull out species that need seqs to be added to the ref db 
```{r}
deacon <- asv_table %>%
  filter(common_name == "deacon") %>%
  select(where(~ !is.numeric(.) || (is.numeric(.) && sum(.) > 0)))
```

interesting.. lots of different asvs... next: look at them in geneious. maybe they are actually very similar? 

are these asvs in deacon unique to deacon or also in other rockfish? 
```{r}
deacon_asvs <- deacon %>%
  select(starts_with("ASV")) %>%
  colnames()
```

```{r}
temp1 <- asv_table %>%
  select(2, all_of(deacon_asvs)) %>%
  group_by(common_name) %>%
  summarise(across(ASV84:ASV9034, sum, na.rm = TRUE)) %>%
  t() %>%
  as.data.frame()

colnames(temp1) <- temp1[1,]  
temp1 <- temp1[-1,-46]  

temp2 <- temp1 %>%
  mutate(across(everything(), as.numeric)) %>%
  mutate(total = rowSums(across(where(is.numeric)))) %>%
  mutate(prop_deacon = deacon/total)
```

```{r}
deacon_filtered <- temp2 %>%
  filter(prop_deacon > 0.9) %>%
  filter(total > 100) %>%
  rownames()
```

subset fasta to be just the deacons
```{r}
# Load the necessary libraries
library(Biostrings)

# Define the input and output file paths
fasta_file <- "/genetics/edna/workdir/sebastes/tissues/untrimmed_20241231/filtered/outputs/myasvs.fasta"
output_file <- "/home/kimberly.ledger/rockfish_mb//outputs/deacon_untrimmed.fasta"

# Read the FASTA file
fasta_seqs <- readDNAStringSet(fasta_file)

# Filter the FASTA sequences based on the ASV headers
deacon_seqs <- fasta_seqs[names(fasta_seqs) %in% deacon_asvs]

# Write the filtered sequences to a new FASTA file
writeXStringSet(deacon_seqs, filepath = output_file)
```

look closer at the ciliatus and variabilis asvs 

# DUSKIES

```{r}
cv <- asv_table %>%
  filter(common_name == "dark dusky" | common_name == "light dusky") %>%
  select(where(~ !is.numeric(.) || (is.numeric(.) && sum(.) > 0)))
```

are these asvs unique by species? 
```{r}
cv_asvs <- cv %>%
  select(starts_with("ASV")) %>%
  colnames()
```

```{r}
temp1 <- asv_table %>%
  select(2, all_of(cv_asvs)) %>%
  group_by(common_name) %>%
  summarise(across(ASV47:ASV244, sum, na.rm = TRUE)) %>%
  t() %>%
  as.data.frame()

colnames(temp1) <- temp1[1,]  
temp1 <- temp1[-1,-45]  

temp2 <- temp1 %>%
  mutate(across(everything(), as.numeric)) %>%
  mutate(total = rowSums(across(where(is.numeric)))) %>%
  mutate(prop_light = `light dusky`/total) %>%
  mutate(prop_dark = `dark dusky`/total)
```

yep, the common asvs for both light and dark duskys overlap. can't differentiate with these primers. 

#SUNSET

pull out species that need seqs to be added to the ref db 
```{r}
sunset <- asv_table %>%
  filter(common_name == "sunset") %>%
  select(where(~ !is.numeric(.) || (is.numeric(.) && sum(.) > 0)))
```

are these asvs in deacon unique to deacon or also in other rockfish? 
```{r}
sunset_asvs <- sunset %>%
  select(starts_with("ASV")) %>%
  colnames()
```

```{r}
temp1 <- asv_table %>%
  select(2, all_of(sunset_asvs)) %>%
  group_by(common_name) %>%
  summarise(across(ASV591:ASV9035, sum, na.rm = TRUE)) %>%
  t() %>%
  as.data.frame()

colnames(temp1) <- temp1[1,]  
temp1 <- temp1[-1,-46]  

temp2 <- temp1 %>%
  mutate(across(everything(), as.numeric)) %>%
  mutate(total = rowSums(across(where(is.numeric)))) %>%
  mutate(prop_sunset = sunset/total)
```

```{r}
sunset_filtered <- temp2 %>%
  filter(prop_sunset > 0.9) %>%
  #filter(total > 100) %>%
  rownames()
```

subset fasta to be just the sunsets
```{r}
# Load the necessary libraries
library(Biostrings)

# Define the input and output file paths
fasta_file <- "/genetics/edna/workdir/sebastes/tissues/untrimmed_20241231/filtered/outputs/myasvs.fasta"
output_file <- "/home/kimberly.ledger/rockfish_mb/outputs/sunset_untrimmed.fasta"

# Read the FASTA file
fasta_seqs <- readDNAStringSet(fasta_file)

# Filter the FASTA sequences based on the ASV headers
sunset_seqs <- fasta_seqs[names(fasta_seqs) %in% sunset_filtered]

# Write the filtered sequences to a new FASTA file
writeXStringSet(sunset_seqs, filepath = output_file)
```


#OLIVE

pull out species that need seqs to be added to the ref db 
```{r}
olive <- asv_table %>%
  filter(common_name == "olive") %>%
  select(where(~ !is.numeric(.) || (is.numeric(.) && sum(.) > 0)))
```

are these asvs in deacon unique to deacon or also in other rockfish? 
```{r}
olive_asvs <- olive %>%
  select(starts_with("ASV")) %>%
  colnames()
```

```{r}
temp1 <- asv_table %>%
  select(2, all_of(olive_asvs)) %>%
  group_by(common_name) %>%
  summarise(across(ASV105:ASV7951, sum, na.rm = TRUE)) %>%
  t() %>%
  as.data.frame()

colnames(temp1) <- temp1[1,]  
temp1 <- temp1[-1,-46]  

temp2 <- temp1 %>%
  mutate(across(everything(), as.numeric)) %>%
  mutate(total = rowSums(across(where(is.numeric)))) %>%
  mutate(prop_olive = olive/total)
```


```{r}
olive_filtered <- temp2 %>%
  filter(prop_olive > 0.9) %>%
  filter(total > 100) %>%
  rownames()
```

subset fasta to be just the olives
```{r}
# Load the necessary libraries
library(Biostrings)

# Define the input and output file paths
fasta_file <- "/genetics/edna/workdir/sebastes/tissues/untrimmed_20241231/filtered/outputs/myasvs.fasta"
output_file <- "/home/kimberly.ledger/rockfish_mb/outputs/olive_untrimmed.fasta"

# Read the FASTA file
fasta_seqs <- readDNAStringSet(fasta_file)

# Filter the FASTA sequences based on the ASV headers
olive_seqs <- fasta_seqs[names(fasta_seqs) %in% olive_asvs]

# Write the filtered sequences to a new FASTA file
writeXStringSet(olive_seqs, filepath = output_file)
```


ABLG11655 - doesn't look like it's actually an olive rf


#gopher

pull out species that need seqs to be added to the ref db 
```{r}
gopher <- asv_table %>%
  filter(common_name == "gopher") %>%
  select(where(~ !is.numeric(.) || (is.numeric(.) && sum(.) > 0)))
```

are these asvs in deacon unique to deacon or also in other rockfish? 
```{r}
gopher_asvs <- gopher %>%
  select(starts_with("ASV")) %>%
  colnames()
```

```{r}
temp1 <- asv_table %>%
  select(2, all_of(gopher_asvs)) %>%
  group_by(common_name) %>%
  summarise(across(ASV286:ASV5534, sum, na.rm = TRUE)) %>%
  t() %>%
  as.data.frame()

colnames(temp1) <- temp1[1,]  
temp1 <- temp1[-1,-46]  

temp2 <- temp1 %>%
  mutate(across(everything(), as.numeric)) %>%
  mutate(total = rowSums(across(where(is.numeric)))) %>%
  mutate(prop_gopher = gopher/total)
```


```{r}
gopher_filtered <- temp2 %>%
  filter(prop_gopher > 0.9) %>%
  filter(total > 100) %>%
  rownames()
```

subset fasta to be just the gophers
```{r}
# Load the necessary libraries
library(Biostrings)

# Define the input and output file paths
fasta_file <- "/genetics/edna/workdir/sebastes/tissues/untrimmed_20241231/filtered/outputs/myasvs.fasta"
output_file <- "/home/kimberly.ledger/rockfish_mb/outputs/gopher_untrimmed.fasta"

# Read the FASTA file
fasta_seqs <- readDNAStringSet(fasta_file)

# Filter the FASTA sequences based on the ASV headers
gopher_seqs <- fasta_seqs[names(fasta_seqs) %in% gopher_asvs]

# Write the filtered sequences to a new FASTA file
writeXStringSet(gopher_seqs, filepath = output_file)
```

