---
title: "oncorhynchus_gadid_customdbs"
author: "Kimberly Ledger"
date: "2024-07-08"
output: html_document
---

```{r load-libraries}
library(tidyverse)
rename <- dplyr::rename
```

input blastn results from custom rockfish database
```{r}
rock <- read.table("/home/kimberly.ledger/rockfish_mb/data/blastn_custom519.txt")  %>%
  filter(V4 > 100) %>% # make sure all retained matches are >100 bp
  select(-V5, -V6, -V7, -V8, -V9, -V10, -V11, -V12, -V13, -V14) %>%  #remove unnecessary columns
  group_by(V1, V2) %>% # group by the sequence key and the full taxonomy to reduce duplicate entries
  unique() %>% # doing that reduced the number of entries
  rename(qseqid=V1, accession_id=V2, perc_id=V3, length=V4) %>% #rename headers
  filter(perc_id > 98)
```

load the id's of accessions
```{r}
id <- read.csv("/home/kimberly.ledger/rockfish_mb/data/rockfish_reference_db_519.csv") %>%
  rename(accession_id = Name)
```

join 
```{r}
rock_id <- rock %>%
  left_join(id) %>%
  rename(species = Organism)
```

## rank the hits by % identity
```{r}
seq_id_diff <- rock_id %>%
  select(-length) %>%
  group_by(qseqid, species) %>%
    mutate(seq_percID = max(perc_id)) %>%
    group_by(qseqid, species, seq_percID) %>%
  summarise(max(seq_percID)) %>% # take just the top hit for each taxon (for each sequence)
  select(-`max(seq_percID)`) %>%
  ungroup() %>%
  group_by(qseqid) %>%
      #mutate(id_rank = rank(desc(seq_percID), ties.method = "min")) %>% # rank the taxonomic hits per sequence by % id
       mutate(top_perc = max(seq_percID)) %>% # designate the highest % id for the best taxonomic hit in each sequence (in some, but not all cases, this is 100%)   %>%
      mutate(diff = top_perc - seq_percID) %>% # calculate the difference between the % identity of the top hit and each subsequent taxonomic hit
      arrange(diff)

seq_id_diff
```

since there are very few basepair difference among species, let's go with species within 0.5% of the top match and top match must be at least 99% 
```{r}
keepers <- seq_id_diff %>%
  filter(diff < 0.5) #%>%
  #filter(top_perc > 99)
```


## so now determine whether ASVs should be assigned to genus, family, order, etc. 
```{r}
singletons <- keepers %>%
  select(qseqid) %>%
  tally() %>%
  filter(n == 1)

# these are the seqs that now have only a single match
singleton_df <- singletons %>%
  left_join(keepers) %>%
  select(-n) %>%
  mutate(taxonomic_level = "species") %>%
  mutate(taxon = species) #%>%
  #select(qseqid, taxon, taxonomic_level)

## Genus-level matches
# remove the singletons from the bigger df 
single_genus <- keepers %>%
  anti_join(singleton_df)%>% # at best, these should be genus-level matches
  left_join(., keepers) %>%
  arrange(qseqid) %>%
  mutate(taxonomic_level = "genus") %>%
  mutate(taxon = "Sebastes") #%>%
  #select(qseqid, taxon, taxonomic_level) %>%
  #unique
```

so as I have the Oncorhynchus genus designation right now, it can be any pacific salmon (or O. mykiss) except for Chinook. 

remove underscore from taxon name 
```{r}
rockfish_df <- singleton_df %>%
  bind_rows(single_genus)
```

```{r}
 rockfish_df_2 <- rockfish_df %>%
   mutate(taxon = ifelse(qseqid == "ASV81", "Sebates_pv", taxon)) #,
#          taxon = ifelse(qseqid == "ASV25", "Sebates_pv", taxon),
#          taxon = ifelse(qseqid == "ASV39", "Sebates_pv", taxon),
#          taxon = ifelse(qseqid == "ASV43", "Sebates_pv", taxon),
#          taxon = ifelse(qseqid == "ASV5", "Sebates_az", taxon),
#          taxon = ifelse(qseqid == "ASV63", "Sebates_az", taxon),
#          taxon = ifelse(qseqid == "ASV69", "Sebates_pv", taxon),
#          taxon = ifelse(qseqid == "ASV80", "Sebates_az", taxon))
```

```{r}
rockfish_df_3 <- rockfish_df_2  %>%
   select(qseqid, taxon) %>%
   unique()
```

```{r}
write.csv(rockfish_df_3, "/home/kimberly.ledger/rockfish_mb/data/taxonomy_custom519.csv")
```








maybe 
```{r}

#oncorhynchus_df$taxon <- gsub("_", " ", oncorhynchus_df$taxon)
#oncorhynchus_df$species <- gsub("_", " ", oncorhynchus_df$species)
#oncorhynchus_df$genus <- "Oncorhynchus"
#oncorhynchus_df$family <- "Salmonidae"
#oncorhynchus_df$order <- "Salmoniformes"
#oncorhynchus_df$class <- "Actinopteri"
#oncorhynchus_df$phylum <- "Chordata"
#oncorhynchus_df$kingdom <- "Eukaryota"

#export 
#write.csv(oncorhynchus_df, "../outputs/20240806/oncorhynchus_asv_id.csv")
```


