---
title: "Taxonomic Assignment"
author: "Kimberly Ledger"
date: "2024-11-06"
output: html_document
---

WARNING!!!! not using this in updated version (Nov 2024) of data analysis 


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load-libraries}
library(tidyverse)
library(tidyr)
library(stringr)
rename <- dplyr::rename
```

## load output from taxonkit 
```{r load-output-from-taxonkit}
taxonomy <-read.delim("/home/kimberly.ledger/rockfish_mb/data/blastn_taxlineage_98.txt", header = FALSE, na.strings=c(""," ","NA"))

head(taxonomy)
```


## clean up the header a bit
```{r}
# use the full taxonomy rather than the seq id to collapse identical entries
tax_df <- taxonomy %>%
  filter(V4 > 100) %>% # make sure all retained matches are >100 bp
  select(-V2, -V5, -V6, -V7, -V14, -V8, -V9, -V10, -V11, -V12, -V13, -V15, -V16) %>%  #remove unnecessary columns
  group_by(V1, V17) %>% # group by the sequence key and the full taxonomy to reduce duplicate entries
  unique() %>% # doing that reduced the number of entries
  rename(qseqid=V1, perc_id=V3, length=V4, taxonomy=V17) %>% #rename headers
  filter(perc_id > 98) %>% 
  filter(!is.na(taxonomy))
```

## calcualate the number of ASV's with taxonomic assignment
```{r how-many-total-ASVs-w-taxonomy}
tax_df %>% 
  ungroup() %>%
  select(qseqid) %>%
  unique() %>%
  nrow
```

## reformat table 
```{r}
# formatting the taxonomy variables
taxon_df <- tax_df %>%
  filter(str_detect(taxonomy, ";")) %>%
  separate(taxonomy, into=c("kingdom", "phylum", "class", "order", "family", "genus", "species"), sep = ";") %>% 
  filter(!grepl("sp\\.", species, ignore.case = TRUE))
```


## figure out what ASVs are unambiguously a single species
```{r}
single_spp_seqs <- taxon_df %>% 
  group_by(qseqid) %>%
  add_tally(name = "n_taxa") %>%
  filter(n_taxa == 1)
```

## remove the single-species seqs from the dataframe and then rank the hits by % identity for the remaining seqs
```{r}
seq_id_diff <- taxon_df %>%
  anti_join(., single_spp_seqs) %>%  ## comment out this line if there are no single species ASVs
  select(-length) %>%
  group_by(qseqid, species, genus, family, order, class, phylum, kingdom) %>%
    mutate(seq_percID = max(perc_id)) %>%
    group_by(qseqid, species, genus, family, order, class, phylum, kingdom, seq_percID) %>%
  summarise(max(seq_percID)) %>% # take just the top hit for each taxon (for each sequence)
  select(-`max(seq_percID)`) %>%
  ungroup() %>%
  group_by(qseqid) %>%
      #mutate(id_rank = rank(desc(seq_percID), ties.method = "min")) %>% # rank the taxonomic hits per sequence by % id
       mutate(top_perc = max(seq_percID)) %>% # designate the highest % id for the best taxonomic hit in each sequence (in some, but not all cases, this is 100%)   %>%
      mutate(diff = top_perc - seq_percID) %>% # calculate the difference between the % identity of the top hit and each subsequent taxonomic hit
        arrange(diff)
```

Now you have the single best entry for each species for each sequence ranked and with the difference between the first and second ranked entries calculated.

For sequences with multiple top hits, where the difference between ranked taxa = 0, you will end up defaulting to genus- or family-level ID. We'll do the same for any sequences where the difference between the first and second ranked taxa is < 1%.

** if there is a sequence that is a 100% match, then keep just the sequences within 0.5%  
** if there is NOT a 100% match, keep the top 1% 


## Figure out which differences are **> 0.5%** and eliminate those first
```{r}
# filter out any taxa that are >1% less matching identity than the top taxonomic hit for a given sequence
low_perc_hits <- seq_id_diff %>%
  ungroup() %>%
  group_by(qseqid) %>%
  filter(diff > 0.5)

keepers <- seq_id_diff %>%
  anti_join(low_perc_hits)
```

## so now determine whether ASVs should be assigned to genus, family, order, etc. 
```{r}
singletons <- keepers %>%
  select(qseqid) %>%
  tally() %>%
  filter(n == 1)

# these are the seqs that now have only a single match
singleton_df <- singletons %>%
  left_join(keepers) %>%
  select(-n) %>%
  bind_rows(single_spp_seqs) %>% # combine the single spp data
  mutate(taxonomic_level = "species") %>%
  mutate(taxon = species)

## Genus-level matches
# remove the singletons from the bigger df 
single_genus <- keepers %>%
  anti_join(singleton_df)%>% # at best, these should be genus-level matches
  group_by(qseqid, genus) %>%
  tally() %>%
  ungroup() %>%
  group_by(qseqid) %>%
  tally() %>%
  filter(n == 1) %>% # seqs that match a single genus
  select(-n) %>%
  left_join(., keepers) %>%
  mutate(taxonomic_level = "genus") %>%
  mutate(taxon = genus)
```

## modify the singleton_df to include the right variable headers
```{r}
single_spp <- singleton_df %>%
  select(-perc_id, -length, -n_taxa) %>%
  mutate(taxonomic_level = "species") %>%
  mutate(taxon = species)
```

## recombine the full data set now that the appropriate level of taxonomy has been determined
```{r}
tax_df_2 <- bind_rows(single_genus, single_spp)
```

## mutate the taxon column to designate which sebastes species are lumped together
```{r}
tax_df_3 <- tax_df_2 %>%
  mutate(taxon = ifelse(qseqid == "ASV13", "Sebates_cpv", taxon),
         taxon = ifelse(qseqid == "ASV20", "Sebates_pv", taxon),
         taxon = ifelse(qseqid == "ASV25", "Sebates_pv", taxon),
         taxon = ifelse(qseqid == "ASV39", "Sebates_pv", taxon),
         taxon = ifelse(qseqid == "ASV42", "Sebates_pv", taxon),
         taxon = ifelse(qseqid == "ASV43", "Sebates_pv", taxon),
         taxon = ifelse(qseqid == "ASV66", "Sebates_em", taxon),
         taxon = ifelse(qseqid == "ASV67", "Sebates_cpv", taxon),
         taxon = ifelse(qseqid == "ASV69", "Sebates_pv", taxon),
         taxon = ifelse(qseqid == "ASV71", "Sebates_pv", taxon),
         taxon = ifelse(qseqid == "ASV87", "Sebates_pv", taxon))
```

```{r}
tax_df_4 <- tax_df_3 %>%
  select(qseqid, taxon) %>%
  unique()
```

```{r}
write.csv(tax_df_3, "/home/kimberly.ledger/rockfish_mb/data/taxonomy_uncollapsed.csv")
write.csv(tax_df_4, "/home/kimberly.ledger/rockfish_mb/data/taxonomy_collapsed.csv")
```

